---
title: "M358K-Applied Statistics"
author: "ArunK"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[`r knitr::asis_output("\U1F332")`](https://arun-krishnaraj.github.io/evergreen)

### Notes
#### Lecture 9/2/2020
- Some introductory R programming review
- For every random variable $X$, its cumulative distribution function is the function $F_{x}: \mathbf{R} \rightarrow [0,1]$ given by $F_x (x) = \mathbf{P}[X \leq x]$ for all $x \space \epsilon \mathbf{R}$

#### 9/4/2020
- The simplest discrete random variable is the Bernoulli trial:
X~ 0 with probability 1-p(=q); 1 with probability p
-Its probability mass function is:
  $p_x(0) = \mathbb{P}[x=0]=1-p$,
  
  $p_x(1) = \mathbb{P}[x=1]=p$

- Note: this is an example of a named distribution with a single parameter $p$ interpreted as the probability of success in a single trial
- What happens if you repeat a number n = # of trials = size, of independent Bernoulli trials each with the same parameter $p$.
Let $X_1,...,X_n$ be independent Bernoulli(p). Then your total # of successes will be $S=X_1+X_2+...+X_n$, which results in Binomial(n,p). For the pmf: $p_S(k)= choose(n k) p^k(1-p)^{n-k})$ fro all k = 0...n.

#### Lecture 9/9/2020
- What is statistics?
  - Identify the broad population of interest about which you're trying to draw conclusions
  - What are we trying to measure?
  - The members of the population which you are trying to measure
    - cases (observational units) should become private labels 
    - variables can be quantitative (discrete or continuous) or qualitative (nominal or ordinal)
  - focus on figuring out the relative likelihoods of different outcomes
    - Probabilistic model:
      - named distribution with input parameters:
        - binomial(n,p); p is usually the parameter of interest 
        - geometric(p)
        - Poisson($\lambda$)
        - Normal(mean = $\mu$, variance = $\sigma^2$)
      - we are often interested in the behavior of non-named distributions 
  - collect and analyze data: find appropriate shape of distribtion and/or values of interest

#### Lecture 9/11/2020
- Using data sets pie-survey and titanic for some exploratory analysis

```{r}
pie <- read.csv("pie-survey-1.csv")
pielist <- table(pie$What.is.your.favorite.kind.of.pie.)
barplot(pielist)
pie(pielist)
```

#### Lecture 9/14/2020
```{r}
cafe <- read.csv("cafedata-1.csv")
plot(cafe$Wraps.Sold)
wraps <- cafe$Wraps.Sold[cafe$Wraps.Sold != "na"]
wraps <- as.numeric(wraps)
hist(wraps)
```

- sample mean, and other sample statistics, serve as point estimations of a true population statistic
  - data visualizations should be meaningful and accurate depictions; modality and skewedness
  - variance is a non-standardized measure of spread from the mean; weighs larger deviations more heavily, removes signs from deviation calculations

#### Lecture 9/16/2020
- median is the value that splits the data in half when ordered (mean of two medians if there are even observations)
  - IQR is the range between Q3 and Q1 (75th and 25th percentiles) and contains 50% of the sample data
    - add or subtract 1.5 times the IQR to Q3 or Q1 for whiskers 
  - more robust than mean to large outliers or deviations; good for highly skewed distributions
- highly skewed data can be transformed to be better (common log transform)
- exploratory base stats in the diamond dataset

#### Lecture 9/21/2020
- randomized response includes a randomized question and response to generate honest samples 
- observational studies: data is collected indirectly from the data arising 
  - prospective vs retrospective: identifies data prior or after events have taken place
- random sampling techniques:
  - simple: every sample is equally likely
  - stratified: split groups to better capture proportion
  - cluster: works if subsets of sample are alike 

### Reading notes

Section 4.3 
- binomial distribution is used to describe the number of successes in a fixed number of trials 
  - describes the probability of exactly k successes in n independent Bernoulli trials with probability of success p
  $(n \text{ choose } k)\ p^k(1-p)^{n-k}, (n \text{ choose } k) = \frac{n!}{k!(n-k)!}$
- the normal distribution can be used to approximate the binomial distribution when the sample size is large
  - requires estimation of a large range of counts, or else performs poorly
  
Section 2.1.3
- plotting individual points can be bad for larger samples, instead bin and display counts in histogram
  - provide clear depiction of data density and distribution
  - skew occurs when tails occur in the direction of choice; roughly equal tails are symmetric 
  - mode: prominent peak in distribution, can have one, two or many modes
  
Section 1.2.1
- observations or cases appear as rows in data, with columns of variables

Section 1.2.2
- variables can be numeric, categorical, and continuous or discrete and nominal or ordinal

Section 2.2.1
- contingency tables summary count data for two categorical variables; individual groups, row, column and grand totals

Section 2.2.5
- pie charts present proportional information on categories

Section 2.1.2
- dot plot is a one variable scatterplot; the mean is the average, common way to measure the center of a distribution

Section 2.1.4
- deviance measures the distance from the mean, the mean of the squared deviation is the variance; sqrt of variance is sd

Section 2.1.5
- box plots summarize outliers, median, Q1 and Q3, upper and lower whiskers and outliers
- median is the observation splitting the data in 5 by counts; whiskers can only reach up to 1.5 times the IQR, never more

Section 2.1.6
- median and IQR are robust since extreme observations have little effect on them, compared to sd and mean

Section 1.2.3
- variables with some connection to each other are called associated, and can be called dependent
  - association can be positive or negative; otherwise they are independent
  
Section 1.3 Sampling principles and strategies
- research questions refer to individual target populations, of which a sample is taken to represent the average population member 
- anecdotal evidence is single cases used to draw general conclusions
- randomly selected samples are best
  - simple random sample is equivalent to raffle
  - non-response rate can influence sample composition
  - convenience sample: easily accessible people are more likely to be included in the sample 
- observational data occurs when no treatment is explicitly applied 
  - prospective studies identify individuals and collect data as it arises, compared to retrospective studies that collect data after events have taken place
- stratified sampling divides samples into similar cases then samples from each one
  - useful when stratum cases are similar in outcome of interest
- cluster sample has population broken into many clusters, with a fixed number of clusters being selected to generate the sample; multistage sampling can apply simple sampling to chosen clusters