---
title: "M358K-Project1"
author: "ArunK"
date: "9/15/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
chuck <- replicate(sum(sample(1:6,1,replace = TRUE) == sample(1:6,3,replace = TRUE)), n =100)
chuck_dol <- ifelse(chuck == 0, -1, ifelse(chuck == 1, 1, ifelse(chuck == 2,2,ifelse(chuck==3,3,0))))
mean(chuck_dol)
```
#### Problem 1.1 
The probability of winning is 1 minus the probability of losing. Losing corresponds to the event where no dice land on the gambler's number, so $ \frac{5}{6}^3= \frac{125}{216}$, and the probability of winning is $1-\frac{125}{216} = \frac{91}{216}$

#### Problem 1.2 
The house has a long run advantage, since the probability of the gambler winning is less than half.

#### Problem 1.3 
We're interested in making the game fair by adjusting the winnings.
$\frac{125}{216}*\$1 = \frac{1}{216}\$b +\frac{15}{216}\$a+\frac{75}{216}\$1 \rightarrow \$50 = \$b + \$15a$
so we could have any nonnegative solution to this 

#### Problem 1.4
average winnings of $.12 during the 100 rounds

```{r}
gamble <- function(incash,outcash) {
cash <- c(incash)
while (cash[length(cash)] > 0 & cash[length(cash)] < outcash) {
  samp <- sum(sample(1:6,3,replace = T) == sample(1:6,1,replace = T))
  if (samp == 0) {
    cash <- c(cash, cash[length(cash)]-1)
  }
  if (samp == 1) {
    cash <- c(cash, cash[length(cash)]+1)
  }
  if (samp == 2) {
    cash <- c(cash, cash[length(cash)]+2)
  }
  if (samp == 3) {
    cash <- c(cash, cash[length(cash)]+3)
  }
  } 
return(cash)
}

replicate(gamble(1,4),n =20)
```

